Unshuffling Permutations

### 2015-09-17 article.tex
Reorganisation intro

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}

Given three words $u$, $v$, and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word. In addition to purely combinatorial questions, the
shuffle product of words leads to the following computational questions:
\begin{enumerate}
    \item \label{item:problem_1}
    Given three words $u$, $v$, and $w$, decide if $u$ is a shuffle of
    $v$ and $w$.
    \item \label{item:problem_2}
    Given a word $u$, decide if there exist two words $v$ and $w$ such
    that $u$ is a shuffle of $v$ and $w$.
\end{enumerate}
Problem~\ref{item:problem_1} is in $\Pclass$ (it is a classical
textbook exercise to design an efficient dynamic programming algorithm
that determines whether $u$ is a shuffle of $v$ and $w$), while
Problem~\ref{item:problem_2} is $\NPclass$-complete as shown
in~\cite{Rizzi:Vialette:CSR:2013}. Analogous questions can be stated to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that
$u$ is a shuffle of $v$ with itself.
\smallskip

This paper is intended to study a natural generalization $\SHUFFLE$
of the shuffle
of permutations. Roughly speaking, given three permutations $\pi$, $\mu$,
and $\nu$, $\pi$ is said to be a {\em shuffle} of $\mu$ and $\nu$ if $\pi$
(viewed as a word) is a shuffle of two words that are order-isomorphic
to $\mu$ and $\nu$.
%For
%example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
%permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
%exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
%itself. For example, $\pi = 241356$ is a square as it is the shuffle of
%$213$ with itself (\emph{e.g.},
%$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
%denote the second occurrence), whereas $\pi = 4123$ is not.
Our intention in this paper is to study this shuffle product of
permutations both from a combinatorial and from a computational point of
view by focusing on {\em square} permutations, that are permutations
$\pi$ being in the shuffle of a permutation $\mu$ with itself. The
shuffle product defined and considered in this paper is different from
other existing and well-known shuffle products on permutations. For
instance, in~\cite{DHT:IJAC:2002}, the authors define the
{\em shifted shuffle product}. For this product, $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is in the shuffle, as words, of $\mu$ and the
word obtained by incrementing all the letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, given three permutations $\pi$,
$\mu$, and $\nu$, deciding if $\pi$ is a in the shifted shuffle of
$\mu$ and $\nu$ is in $\Pclass$.
\smallskip

Let us now  briefly review the key results for words that arise in our context.
%Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
%$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
%where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
%such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
%first briefly review the key results for words that arise in our context.
Given three words $u$, $v_1$, and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u$ is in the shuffle $v_1 \shuffle v_2$ of $v_1$ and
$v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
To the best of our knowledge, the first $O(|u|^2)$ time algorithm
appeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words. The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
\begin{equation}
    O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1$, $u_2$, \dots, $u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
\begin{equation}
    O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)
\end{equation}
time. Given words $u$, $v_1$, $v_2$, \dots, $v_n$, it is however \NPC
to decide whether or not $u$ is in $\shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the ground alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the ground alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a precise
definition of $\SHUFFLE$. This definition passes
through the preliminary definition of an operator $\Delta$ to unshuffle
permutations. This operator is in fact a coproduct, endowing the space
of all permutations with a coalgebra structure. By duality, the
unshuffling operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some combinatorial
properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct
side. Second, this way of doing allows to obtain a clear description of
the multiplicities of the elements appearing in the shuffle of two
permutations, which are worthy of interest.
Section~\ref{section:Binary square words and permutations}
is devoted to show that the problems related to the shuffle of words has
links with the shuffle of permutations. In particular, we show that
binary words that are square are in one-to-one bijection with square
permutations (Proposition~\ref{prop:bijection_binary_to_permutations_squares}).
Next, Section~\ref{section:Algebraic issues} presents some algebraic
and combinatorial properties of~$\SHUFFLE$. In particular, we show that
$\SHUFFLE$ is associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that
if a permutation if a square, its mirror, complement, and inverse also
are square (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents
the most import result of this paper: the fact that deciding if
a permutation is a square is $\NPclass$-complete
(Proposition~\ref{proposition:hardness}).

$123$-avoiding (resp. $321$-avoiding) permutations are the union of two decreasing
(resp. increasing) patterns.

>>>>

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and MacLane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v$, and $w$, $u$ is said to be a \emph{shuffle}
of $v$ and $w$ if it can be formed by interleaving the letters from $v$
and $w$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there exist two words $v_1$ and $v_2$
    such that $u$ is a shuffle of $v_1$ and $v_2$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that, given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. The hardness of this last particular
problem implies that Problem~\ref{item:problem_4} is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\mu$, and $\nu$, $\pi$ is said to be a
{\em shuffle} of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is a
shuffle of two words that are order-isomorphic to $\mu$ and $\nu$.
Our intention in this paper is to study this shuffle product of
permutations $\SHUFFLE$ both from a combinatorial and from a
computational point of view by focusing on {\em square} permutations,
that are permutations $\pi$ being in the shuffle of a permutation $\mu$
with itself. The shuffle product defined and considered in this paper
is different from other existing and well-known shuffle products on
permutations. For instance, in~\cite{DHT:IJAC:2002}, the authors define
the {\em shifted shuffle product}. For this product, $\pi$ is a shuffle
of $\mu$ and $\nu$ if $\pi$ is in the shuffle, as words, of $\mu$ and
the word obtained by incrementing all the letters of $\nu$ by the size
of $\mu$. It is a simple exercise to prove that, given three
permutations $\pi$, $\mu$, and $\nu$, deciding if $\pi$ is a in the
shifted shuffle of $\mu$ and $\nu$ is in~\Pclass.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator $\Delta$ to unshuffle permutations.
This operator is in fact a coproduct, endowing the space of all
permutations with a coalgebra structure. By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest. Section~\ref{section:Binary square words and permutations}
is devoted to show that the problems related to the shuffle of words has
links with the shuffle of permutations. In particular, we show that
binary words that are square are in one-to-one bijection with square
permutations
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. In particular, we show that
$\SHUFFLE$ is associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that
if a permutation if a square, its mirror, complement, and inverse also
are square (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents
the most import result of this paper: the fact that deciding if a
permutation is a square is \NPC (Proposition~\ref{proposition:hardness}).
